---
bibliography: References.bib
---

# Analyzing mass spectrometry data in R <br>
## Introduction <br>
Since my last study I have become very interested in microbiology. This has led me to choose to follow my specialization in microbiology at my current study Life Sciences. This has made me even more enthusiastic about the profession. In 2 years time I see myself working in diagnostics at a medical microbiology laboratory. To achieve this goal, I plan to complete my graduation internship in a similar direction as well. <br>

In microbiology, MALDI-TOF MS is currently a widely used method to quickly identify a bacterial species. For this reason, I thought it would be interesting to process raw data from a mass spectrometry experiment in R. <br>

## Required packages <br>
The following packages must be installed to perform the code. If you want to use this code I highly recommend to download the required packages before running the code. <br>
```
# install.packages("tidyverse")
# install.packages("ggplot2")
# install.packages("Biocmanager")
# install.packages("pheatmap")
# Biocmanager::install("msdata")
# BiocManager::install("Spectra")
# BiocManager::install("rpx")
# BiocManager::install("mzR")
# BiocManager::install("PSMatch")
```

## Data import
To analyze mass spectrometry (MS) data I used the rpx package, which consists of multiple MS data files. The data in this package is linked to the proteomexchange database. <br>
Link: <br>
https://proteomecentral.proteomexchange.org/cgi/GetDataset <br>

```{r import data, warning=FALSE, message=FALSE}
# Load library
library(rpx)

# Proteomics raw data
## Type "no" if R asks you to create a directory
px <- PXDataset("PXD000001")

# Load data
px

# See all files
pxfiles(px)
```

When the data is loaded into R, we can then download the data files with the pxget function. <br>
```{r download data, warning=FALSE, message=FALSE}
# Retrieve raw data file
fn <- "TMT_Erwinia_1uLSike_Top10HCD_isol2_45stepped_60min_01-20141210.mzML"

# Download file
mzf <- pxget(px, fn)
```

## Spectra class <br>
Next I used the Spectra package as an abstraction to raw mass spectrometry data. Data abstraction, in general, is the process of removing characteristics from something to reduce it to a set of essential elements. To this end, data abstraction creates a simplified representation of the underlying data, while hiding its complexities and associated operations. [@WhatDataAbstraction] <br>
```{r spectra class, warning=FALSE, message=FALSE}
# Load library
library(Spectra)

# Making a Spectra object of the downloaded data
spectra_object <- Spectra(mzf)

# Check
spectra_object
```

After making the spectra object, it is possible to plot the MS1 spectrum of interest and highlight all the peaks that will be selected for MS2 analysis. <br>
```{r plotting spectra object, warning=FALSE, message=FALSE}
# Extracting the MS1 results
ms_2 <- filterPrecursorScan(spectra_object, 2807)

# Check
ms_2

# Plotting the MS1 results
plotSpectra(spectra_object[2807], xlim = c(400, 1000))
abline(v = precursorMz(ms_2)[-1], col = "grey")
abline(v = precursorMz(ms_2)[2], col = "blue")
```

You can filter the MS2 level spectra and find any 2 MS2 spectra that have matching precursor peaks based on the precursor m/z values. This can be visualized using the plotSpectraOverlay and plotSpectraMirror functions. <br>
```{r matching, warning=FALSE, message=FALSE}
# Check overlay
sp2 <- filterMsLevel(spectra_object, 2L)
anyDuplicated(precursorMz(filterMsLevel(spectra_object, 2)))

i <- which(precursorMz(sp2) == precursorMz(sp2)[37])
sp2i <- sp2[i]

# Visualisation of the matching peaks
plotSpectraOverlay(sp2i, col = c("green", "red"))
```

## Backends <br>
With backends it is possible to store mass spectrometry data while providing via the Spectra class a unified interface to use that data. [@fischerMzRParserNetCDF2023] <br>
```{r backends, warning=FALSE, message=FALSE}
# Load library
library(mzR)

# Make backend
## I used the the MsBackendDataFrame option from the mzR package to store the spectra data in a dataframe
setBackend(spectra_object, MsBackendDataFrame())
```

## Checking the data
When the data is stored, we can check the data frame. We want to check the data frame on number of protein matches and the amount of sequences.
```{r inspect dataframe, message=FALSE, warning=FALSE}
# Load library
library(PSMatch)

# Check if your data frame is indeed a data frame
identification <- msdata::ident(full.names = TRUE)
basename(identification)

# Check the identification data
id <- PSM(identification)

names(id)

# Check the amount of peptide sequences
## You can do this for all the shown identification data
length(unique(id$sequence)) 

# Check if forward and reverse databases are present
table(id$isDecoy)
```

```{r matches, warning=FALSE, message=FALSE}
# Number of matches in the data
table(table(id$spectrumID))

# Check matches for specific scan
matches <- which(id$spectrumID == "controllerType=0 controllerNumber=1 scan=1774")
data.frame(id[matches, ])[1:5]
```

Scan 1774 has 4 matches for 4 different proteins. We want to store the data in such a way that the different matches are stored in different rows in the data. <br>
```{r store matches, warning=FALSE, message=FALSE}
id2 <- reducePSMs(id, id$spectrumID)

# Create object 
match_object <- which(id2$spectrumID == "controllerType=0 controllerNumber=1 scan=1774")
id2[match_object, ]

# Identification table
id2[match_object, "DatabaseAccess"]
```

```{r tibble, warning=FALSE, message=FALSE}
# Load library
library(dplyr)

# Make tibble
id_tbl <- as_tibble(id)

# Check
id_tbl

# Delete decoy hits
id_tbl <- id_tbl %>%
    filter(!isDecoy)

# Filter the first rank matches
id_tbl <- id_tbl %>%
    filter(rank == 1)

# Check
id_tbl
```

It is possible to visualize the number of decoy and non-decoy hits. This way you can compare the distribution of the raw identification scores. <br>
```{r decoy, warning=FALSE, message=FALSE}
# Load library
library(ggplot2)

# Make plot
as_tibble(id) %>%
    ggplot(aes(x = MS.GF.RawScore,
               colour = isDecoy)) +
    geom_density()
```

The describePeptides and describeProteins functions from the PSMatch package can provide useful summaries of peptides and proteins in a PSM search result. This can be useful when you are interested in a specific protein. <br>
```{r summaries, warning=FALSE, message=FALSE}
# Number of unique and shared peptides
peptides <- describePeptides(id)

peptides

# Number of proteins defined by only unique, only shared, or a mixture of unique/shared peptides
proteins <- describeProteins(id)

proteins
```

## Merging identification data and raw data <br>
Identification data can be merged into a Spectra object by adding new spectra variables to the appropriate MS2 spectra.
```{r merge, warning=FALSE, message=FALSE}
# Filter PSMs
id_filtered <- filterPSMs(id)

# Reduce PSM table
id_filtered <- reducePSMs(id_filtered, id_filtered$spectrumID)

# Join raw data to identification data
joined <- joinSpectraData(spectra_object, id_filtered,
                      by.x = "spectrumId",
                      by.y = "spectrumID")

# Check variables
spectraVariables(joined)
```

## Visualizing the data <br>
After the data is merged, the data can be visualized based on the identification of a protein or species. <br>

### Countidentifications function
Identification can be established by the countIdentifications function. The function takes a Spectra object containing identification results as input. It then counts the number of identifications each scan (or their descendants) has lead to. This is either 0 or 1 for MS2 scans, or, for MS1 scans, the number of MS2 scans originating from any MS1 peak that lead to an identification. [@CountIdentificationsCountNumber] <br>
```{r visualisation, warning=FALSE, message=FALSE}
# Count identifications
count_idf <- countIdentifications(joined)

# View count identifications
table(msLevel(count_idf), count_idf$countIdentifications)

## Below, we see on the second line that 3457 MS2 scans lead to no PSM, while 2646 lead to 1 identification. Thirthy MS1 scans generated one MS2 scan that lead to a PSM.

# Visualization of the total ions
count_idf |>
filterMsLevel(1) |>
spectraData() |>
as_tibble() |>
ggplot(aes(x = rtime,
           y = totIonCurrent)) +
    geom_line(alpha = 0.25) +
    geom_point(aes(colour = ifelse(countIdentifications == 0,
                                   NA, countIdentifications)),
               size = 0.75,
               alpha = 0.5) +
    labs(colour = "Number of ids")
```

### Comparing spectra <br>
For example, when you use a MALDI-TOF MS, the system compares the spectra of the organism with a database. This can be mimicked in R with the compareSpectra function. Because I do not have a database of this kind, I will create a new spectra object to mimic the system. The similarity can be visualized in a heatmap. <br>
```{r comparing spectra, warning=FALSE, message=FALSE}
# Load library
library(pheatmap)

# Make a dummy spectra object
dummy <- which(count_idf$sequence %in% c("SQILQQAGTSVLSQANQVPQTVLSLLR", "TKGLNVMQNLLTAHPDVQAVFAQNDEMALGALR"))

dummy_idf <- count_idf[dummy]

# Check similarity 
matrix <- compareSpectra(dummy_idf)
rownames(matrix) <- colnames(matrix) <- strtrim(dummy_idf$sequence, 2)

# Plot the similarity
pheatmap::pheatmap(matrix)
```