# Parameterized report for the COVID-19 cases <br>
## Introduction <br>
In this report I will be analyzing the number of COVID-19 cases and deaths from 2020 till 2022. In this analysis I will be using parameters to visualize the number of COVID-19 cases and deaths. To execute this report I downloaded a dataset from the following website: <br>
https://www.ecdc.europa.eu/en/publications-data/data-daily-new-cases-covid-19-eueea-country <br>

## Required packages <br>
The following packages must be installed to perform the code. If you want to use this code I highly recommend to download the required packages before running the code. <br>
```{r library, warning=FALSE, message=FALSE}
# install.packages("tidyverse")
# install.packages("ggplot2")
# install.packages("dplyr")
```

## Importing and filtering the data <br>
For this report the data from Austria will be filtered to visualize the cases and deaths in 2022. 
```{r import and filter, warning=FALSE, message=FALSE}
# Load required libraries
library(dplyr)

# Import data from raw data folder
covid_data <- read.csv(file = "./Data_raw/data.csv")

# Set parameters
# You can change these parameters to any you like
country <- "Austria"
year <- 2022
period <- 10

# Filter the data based on parameters
filtered_data <- covid_data %>%
  filter(countriesAndTerritories == country,
         year == year,
         month == period)
```


## Data visualisation <br>
```{r data visualisation, warning=FALSE, message=FALSE}
# Load library
library(ggplot2)

# Generate the graph for COVID-19 cases
cases_plot <- ggplot(filtered_data, aes(x = as.Date(paste(year, month, day, sep = "-")), y = cases)) +
  geom_line() +
  labs(x = "Date", y = "Number of Cases", title = "Number of COVID-19 cases") +
  theme_minimal()

# See plot cases
cases_plot

# Generate the graph for COVID-19 deaths
deaths_plot <- ggplot(filtered_data, aes(x = as.Date(paste(year, month, day, sep = "-")), y = deaths)) +
  geom_line() +
  labs(x = "Date", y = "Number of Deaths", title = "Number of deaths caused by COVID-19 infection") +
  theme_minimal()

# See plot deaths
deaths_plot
```